---
title: "HW 2: CONJ 620 W22"
author: "Your name here - update this!!!!"
date: "Due Wed 1/19/22"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 5)
```


```{r message = FALSE}
library(tidyverse)
library(janitor)
```

## Directions

* Please upload your homework to Sakai. Upload both your .Rmd code file and the knitted .html file.
    + If you do not upload the html file of your knitted Rmd file, 20 (out of 100) points will be deducted from your homework grade. _If there is an error with code in a code chunk that is preventing the file from knitting, then comment out that code and make a note that it's not running._ 
* The questions are numbered 1-14. Please do not use `(@)` in your edits to this file so that the exercise numbering does not change!
* For each question, make sure to include all code and resulting output in the html file to support your answers. Answers to questions do not need be lengthy, but thoughtful and descriptive. 

> Tip: It is a good idea to try kitting your document from time to time as you go along!  Note that knitting automatically saves your Rmd file and knitting frequently helps you catch your errors more quickly.

# Hypothyroidism data background

* The data are from a random sample of patients from the study described below.
* _Effects of Altering Levothyroxine Dose on Energy Expenditure and Body Composition in Subjects Treated With LT4_ by Samuels et al. (J Clin Endocrinol Metab, November 2018, 103(11):4163–4175). 

From the paper:

> __Background:__ It is unclear whether variations in thyroid status within or near the reference range affect energy expenditure, body mass, or body composition.

> __Methods:__ 138 subjects treated with levothyroxine (LT4) for hypothyroidism with normal TSH levels underwent measurement of total, resting, and physical activity energy expenditure; thermic effect of food; substrate oxidation; dietary intake; and body composition. They were assigned to receive an unchanged, higher, or lower LT4 dose in randomized, double-blind fashion, targeting one of three TSH ranges (0.34 to 2.50, 2.51 to 5.60, or 5.61 to 12.0 mU/L). The doses were adjusted every 6 weeks to achieve target TSH levels. Baseline measures were reassessed at 6 months.

# Load the data

The data are in the file `tsh_hw2.csv`.


Explore the dataset. 

(@) __<font color="green">How many observations and variables are included in this dataset? Are there repeated measures amongst the patients? How did you determine this?</font>__



# Data wrangling

(@) __<font color="green">Make a table of the values in the `baseline_group` group variable. Add a new variable to the dataset called `TSH_group_baseline`, which renames the levels of `baseline_group` so that 1 is now `LowNormal` and 2 is `HighNormal`.</font>__



(@) __<font color="green">Create a new dataset that only includes the variables `id`, `time`, `TSH_group_baseline`, and `tsh`. Check that your code did what you expected it to.</font>__





## Make data wide

* In this section we are going to use two different ways to make the long (by time point) data wide, using the smaller dataset created in the previous problem.
* The goal is to create a dataset with separate columns for the baseline and endpoint TSH values.

### Method 1

(@) __<font color="green">Create two separate datasets, where one is restricted to the baseline values and the other is restricted to the endpoint values. Rename the `tsh` variable in the two datasets as `tsh_baseline` and  `tsh_endpoint`, respectively. Check that your code did what you expected it to.</font>__



(@) __<font color="green">Remove the `time` column from the two datasets. Check that your code did what you expected it to.</font>__



(@) __<font color="green">Merge the baseline data and endpoint data together as one dataset with separate columns for the baseline and endpoint TSH data. Check that your code did what you expected it to.</font>__



### Method 2

(@) __<font color="green">Use one of the `pivot` commands to make the dataset wide for the TSH values at the two different time points. Check that your code did what you expected it to. </font>__



# TSH values: baseline vs. endpoint

(@) __<font color="green">Make a scatterplot with endpoint TSH values on the vertical axis, baseline TSH values on the horizontal axis, with dots color coded by the  `TSH_group_baseline` variable.</font>__



(@) __<font color="green">Describe the association between the baseline and endpoint TSH values. </font>__



(@) __<font color="green">Add a variable called `TSH_group_endpoint` to the dataset categorizing the endpoint TSH values as being in the TSH groups listed below:  </font>__
<font color="green">
* `LowNormal` TSH (0.34 to 2.50 mU/L)
* `HighNormal` TSH (2.51 to 5.60 mU/L)
* `Elevated` TSH (greater than 5.60 mU/L)
</font>




(@) __<font color="green">Calculate minimum and maximum summary statistics for the endpoint TSH values stratified by `TSH_group_endpoint`. Do the min and max values agree with the group values listed above?</font>__




(@) __<font color="green">Make a contingency table of the TSH baseline vs. endpoint groups, that includes row and column totals.</font>__



(@) __<font color="green">What percent of people that started in the high normal TSH group ended in the low normal TSH group? </font>__


(@) __<font color="green">What percent of people that started in the low normal TSH group ended in the elevated TSH group? </font>__



